Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-06 talk/icooolps2009/paper.tex
A round of language review.

\begin{abstract}
We present techniques for improving the results when a tracing JIT compiler is
applied to an interpreter. An unmodified tracing JIT performs not as well as one
would hope when the compiled program is itself a bytecode interpreter. We
examine why that is the case, and how matters can be improved by adding hints to
the interpreter, that help the tracing JIT to improve the results. We evaluate
the techniques by using them both on a small example as well as on a full Python
interpreter. This work has been done in the context of the PyPy project.

@@

Dynamic languages have seen a steady rise in popularity in recent years.
JavaScript is increasingly being used to implement full-scale applications
running in browser, whereas other dynamic languages (such as Ruby, Perl, Python,
PHP) are used for the server side of many web sites, as well as in areas
unrelated to the web.

@@

In this paper we discuss ongoing work in the PyPy project to improve the
performance of interpreters written with the help of the PyPy toolchain. The
approach is that of a tracing JIT compiler. Opposed to the tracing JITs for dynamic
languages that exist so far, PyPy's tracing JIT operates "one level down",
e.g. traces the execution of the interpreter, as opposed to the execution
of the user program. The fact that the program the tracing JIT compiles is
in our case always an interpreter brings its own set of problems. We describe
tracing JITs and their application to interpreters in Section
\ref{sect:tracing}.  By this approach we hope to get a JIT compiler that can be
applied to a variety of dynamic languages, given an interpreter for them. The
process is not completely automatic but needs a small number of hints from the
interpreter author, to help the tracing JIT. The details of how the process
integrates into the rest of PyPy will be explained in Section
\ref{sect:implementation}. This work is not finished, but already produces some
promising results, which we will discuss in Section \ref{sect:evaluation}.

@@

\arigo{We should not start from scratch and insert as little details what
differs in our approach when compared to others; instead we should give a
higher-level overview and then focus on these details, and a couple of
references for more info about the "common" part.
%
In general there are many things that are never said at all.
I think the introduction should really be written from the point of view of
someone that has read already some papers for JavaScript.}

@@

At first, when the program starts, everything is interpreted.
The interpreter does a bit of lightweight profiling to figure out which loops
are run often. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
was executed. Since loops need a backward jump somewhere, this method finds
loops in the user program.

@@

Such a history is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by emitting
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code. To ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example conditions or indirect/virtual
calls.  When emitting the machine code, every guard is turned into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit from the machine code and continue the execution by falling
back to interpretation.

During tracing, the trace is repeatedly
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the interpreted program. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
starting from the next iteration of the loop, as the machine code represents
exactly the loop that was being interpreted so far.

This process assumes that the path through the loop that was traced is a
"typical" example of possible paths (which is statistically likely). Of course
it is possible that later another path through the loop is taken, in which case
one of the guards that were put into the machine code will fail. There are more
complex mechanisms in place to still produce more code for the cases of guard
failures \cite{XXX}, but they are orthogonal to the issues discussed in this
paper.

@@

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}
\vspace{-0.4cm}
To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed and Python-specifics have been made explicit. At
first those functions will be interpreted, but after a while, profiling shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often.  The
tracing JIT will then start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
\vspace{-0.4cm}
The operations in this sequence are operations of the mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed, is in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its own beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} was inlined into the trace. Note that the trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be turned into machine code and executed.

@@

The tracing JIT of the PyPy project is atypical in that it is not applied to the
user program, but to the interpreter running the user program. In this section
we will explore what problems this brings, and how to solve them (at least
partially). This means that there are two interpreters involved, and we need
terminology to distinguish them. On the one hand, there is the interpreter that
the tracing JIT uses to perform tracing. This we will call the \emph{tracing
interpreter}. On the other hand, there is the interpreter that is running the
user's programs, which we will call the \emph{language interpreter}. In the
following, we will assume that the language interpreter is bytecode-based. The
program that the language interpreter executes we will call the \emph{user
program} (from the point of view of a VM author, the "user" is a programmer
using the VM).

@@

\fijal{I find following paragraph out of scope and completely confusing, we
should instead simply state that we unroll the loop, how we do that and
why we do that. Completely ignore aspect of an interpreter loop I suppose,
because everything previously keeps talking about can\_enter\_jit that closes
loop being available at jump back bytecodes}
A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot loop
of the language interpreter is its bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\vspace{-0.4cm}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

\fijal{This paragraph should go away as well}
Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers. A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

The program counter of the language interpreter can only be the same a
second time after an instruction in the user program sets it to an earlier
value. This happens only at backward jumps in the language interpreter. That
means that the tracing interpreter needs to check for a closed loop only when it
encounters a backward jump in the language interpreter. Again the tracing JIT
cannot known which part of the language interpreter implements backward jumps,
so it needs to be told with the help of a hint by the author of the language
interpreter.

The condition for reusing already existing machine code needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. They correspond to different
unrollings and paths through that loop though. To figure out which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be entered. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

In addition to the classification of the variables, there are two methods of
\texttt{JitDriver} that need to be called. Both of them get as arguments the
current values of the variables listed in the definition of the driver. The
first one is \texttt{jit\_merge\_point} which needs to be put at the beginning
of the body of the bytecode dispatch loop. The other, more interesting one, is
\texttt{can\_enter\_jit}. This method needs to be called at the end of any
instruction that can set the program counter of the language interpreter to an
earlier value. For the example this is only the \texttt{JUMP\_IF\_A}
instruction, and only if it is actually a backward jump. The place where this
method is called is where the language interpreter performs profiling to decide
when to start tracing. It is also the place where the tracing JIT checks
whether a loop is closed. This is considered to be the case when the values of
the "green" variables are the same as at an earlier call to the
\texttt{can\_enter\_jit} method.

For the small example the hints look like a lot of work. However, the amount of
hints is essentially constant no matter how large the interpreter is, which
makes it seem less significant for larger interpreters.

@@

The critical problem of tracing the execution of just one opcode has been
solved, the loop corresponds exactly to the loop in the square function.
However, the resulting trace is a bit too long. Most of its operations are not
actually doing any computation that is part of the square function. Instead,
they manipulate the data structures of the language interpreter. While this is
to be expected, given that the tracing interpreter looks at the execution of the
language interpreter, it would still be nicer if some of these operations could
be removed.

@@

With this optimization enabled, the trace looks as in Figure
\ref{fig:trace-full}. Now a lot of the language interpreter is actually gone
from the trace and what is left corresponds very closely to the loop of the
square function. The only vestige of the language interpreter is the fact that
the register list is still used to store the state of the computation. This
could be removed by some other optimization, but is maybe not really all that
bad anyway (in fact we have an experimental optimization that does exactly that,
but it is not finished).  Once we get this optimized trace, we can pass it to
the \emph{JIT backend}, which generates the correspondent machine code.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It should be possible to choose when the language interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
as if no hints were present in the interpreter at all.

@@

What is done instead is that the language interpreter keeps running as a C
program, until a hot loop in the user program is found. To identify loops the
C version of the language interpreter is generated in such a way that at the
place that corresponds to the \texttt{can\_enter\_jit} hint profiling is
performed using the program counter of the language interpreter. Apart from this
bit of profiling, the language interpreter behaves in just the same way as
without a JIT.

@@

This section will hint at some other implementation issues and optimizations
that we have done that are beyond the scope of this paper (and will be subject
of a later publication).

@@

\textbf{Allocation Removal:} A key optimization for making the approach
produce good code for more complex dynamic language is to perform escape
analysis on the loop operation after tracing has been performed. In this way all
objects that are allocated during the loop and don't actually escape the loop do
not need to be allocated on the heap at all but can be exploded into their
respective fields.  This is very helpful for dynamic languages where primitive
types are often boxed, as the constant allocation of intermediate results is
very costly.

@@

In this section we try to evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1GiB RAM, using Linux 2.6.27. All benchmarks where
run 50 times, each in a newly started process. The first run was ignored. The
final numbers were reached by computing the average of all other runs, the
confidence intervals were computed using a 95\% confidence level. All times
include the running of the tracer and machine code production to measure how
costly those are.

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
the square of 46340 (the smallest number so that the square still fits into a 32
bit word) using the bytecode of Figure \ref{fig:square}. The results for various
constellations are as follows:

@@

\textbf{Benchmark 4:} Same as before, but with constant folding enabled. This corresponds to the
trace in Figure \ref{fig:trace-full}. This speeds up the square function nicely,
making it about six times faster than the pure interpreter.

@@

Classical partial evaluation has failed to be useful for dynamic language for
much the same reasons why ahead-of-time compilers cannot compile them to
efficient code. If the partial evaluator knows only the program it simply does
not have enough information to produce good code. Therefore some work has been
done to do partial evaluation at runtime. One of the earliest works on runtime
specialisation is Tempo for C \cite{consel_general_1996, consel_uniform_1996}.
However, it is essentially a normal
partial evaluator ``packaged as a library''; decisions about what can be
specialised and how are pre-determined. Another work in this direction is DyC
\cite{grant_dyc:expressive_2000}, another runtime specializer for C. Both of these projects
have a similar problem as DynamoRIO.  Targeting the C language makes
higher-level specialisation difficult (e.g.\ \texttt{malloc} can not be
optimized).

There has been some attempts to do \emph{dynamic partial evaluation}, which is
partial evaluation that defers partial evaluation completely to runtime
to make partial evaluation more useful for dynamic languages. This concept was
introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
small dynamic language based on lambda-calculus. There is some work by one of
the authors to implement a dynamic partial evaluator for Prolog
\cite{carl_friedrich_bolz_automatic_2008}. There are also experiments within the
PyPy project to use dynamic partial evaluation for automatically generating JIT
compilers out of interpreters \cite{armin_rigo_jit_2007}. So far those have not been as
successful as we would like and it seems likely that they will be supplanted
with the work on tracing JITs described here.

@@

We have shown techniques for making it practical to apply a tracing
JIT to an interpreter. Our first benchmarks indicate that these techniques work
really well on small interpreters and first experiments with PyPy's Python
interpreter make it seems likely that they can be scaled up to realistic
examples.

Of course there is a lot of work still left to do. Various optimizations are not
quite finished. Both tracing and leaving machine code is very slow due to a
double interpretation overhead and we might need techniques for improving those.
Furthermore we need to apply the JIT to the various interpreters that are
written with PyPy to evaluate how widely applicable the described techniques
are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
implementation \cite{bolz_back_2008}, a Prolog interpreter or PyGirl, a Gameboy
emulator \cite{XXX}; but also less immediately obvious ones, like Python's
regular expression engine.

>>>>

\begin{abstract}
We present techniques for improving the results when a tracing JIT compiler is
applied to an interpreter. An unmodified tracing JIT performs not as well as one
would hope when the compiled program is itself a bytecode interpreter. We
examine why that is the case, and how matters can be improved by adding markers to
the interpreter, that help the tracing JIT to improve the results. We evaluate
the techniques by using them both on a small example as well as on a full Python
interpreter. This work has been done in the context of the PyPy project.

@@

Dynamic languages have seen a steady rise in popularity in recent years.
JavaScript is increasingly being used to implement full-scale applications
which run within a browser, whereas other dynamic languages (such as Ruby, Perl, Python,
PHP) are used for the server side of many web sites, as well as in areas
unrelated to the web.

@@

In this paper we discuss ongoing work in the PyPy project to improve the
performance of interpreters written with the help of the PyPy toolchain. The
approach is that of a tracing JIT compiler. Contrary to the tracing JITs for dynamic
languages that currently exist, PyPy's tracing JIT operates "one level down",
e.g. it traces the execution of the interpreter, as opposed to the execution
of the user program. The fact that the program the tracing JIT compiles is
in our case always an interpreter brings its own set of problems. We describe
tracing JITs and their application to interpreters in Section
\ref{sect:tracing}.  By this approach we hope to arrive at a JIT compiler that can be
applied to a variety of dynamic languages, given an appropriate interpreter for each of them. The
process is not completely automatic but needs a small number of hints from the
interpreter author, to help the tracing JIT. The details of how the process
integrates into the rest of PyPy will be explained in Section
\ref{sect:implementation}. This work is not finished, but has already produced some
promising results, which we will discuss in Section \ref{sect:evaluation}.

@@

@@

At first, when the program starts, everything is interpreted.
The interpreter does a small amount of lightweight profiling to establish which loops
are run most frequently. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
was executed. Since loops need a backward jump somewhere, this method looks for
loops in the user program.

@@

Such a history is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by generating
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code. To ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example conditions or indirect/virtual
calls.  When generating the machine code, every guard is turned into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit the machine code and continue the execution by falling
back to interpretation.

During tracing, the trace is repeatedly
checked as to whether the interpreter is at a position in the program where it had been
earlier. If this happens, the trace recorded corresponds to a loop
in the interpreted program. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be executed immediately,
starting from the next iteration of the loop, as the machine code represents
exactly the loop that was being interpreted so far.

This process assumes that the path through the loop that was traced is a
"typical" example of possible paths (which is statistically likely). Of course
it is possible that later another path through the loop is taken, in which case
one of the guards that were put into the machine code will fail. There are more
complex mechanisms in place to still produce more code for the cases of guard
failures \cite{XXX}, but they are independent of the issues discussed in this
paper.

@@

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}
\vspace{-0.4cm}
To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed and Python-specifics have been made explicit. At
first those functions will be interpreted, but after a while, profiling shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often  The
tracing JIT will then start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
\vspace{-0.4cm}
The operations in this sequence are operations of the above-mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed, is in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its own beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} was inlined into the trace. Note that the trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

The tracing JIT of the PyPy project is atypical in that it is not applied to the
user program, but to the interpreter running the user program. In this section
we will explore what problems this brings, and suggest how to solve them (at least
partially). This means that there are two interpreters involved, and we need appropriate
terminology to distinguish beween them. On the one hand, there is the interpreter that
the tracing JIT uses to perform tracing. This we will call the \emph{tracing
interpreter}. On the other hand, there is the interpreter that runs the
user's programs, which we will call the \emph{language interpreter}. In the
following, we will assume that the language interpreter is bytecode-based. The
program that the language interpreter executes we will call the \emph{user
program} (from the point of view of a VM author, the "user" is a programmer
using the VM).

@@

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot loop
of the language interpreter is its bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\vspace{-0.4cm}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers. A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

The program counter of the language interpreter can only be the same a
second time after an instruction in the user program sets it to an earlier
value. This happens only at backward jumps in the language interpreter. That
means that the tracing interpreter needs to check for a closed loop only when it
encounters a backward jump in the language interpreter. Again the tracing JIT
cannot know which part of the language interpreter implements backward jumps,
so the author of the language interpreter needs to indicate this with the help
of a hint.

The condition for reusing already existing machine code needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. However, they correspond to different
paths through the loop and different ways to unroll it. To ascertain which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be entered. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

In addition to the classification of the variables, there are two methods of
\texttt{JitDriver} that need to be called. Both of them receive as arguments the
current values of the variables listed in the definition of the driver. The
first one is \texttt{jit\_merge\_point} which needs to be put at the beginning
of the body of the bytecode dispatch loop. The other, more interesting one, is
\texttt{can\_enter\_jit}. This method needs to be called at the end of any
instruction that can set the program counter of the language interpreter to an
earlier value. For the example this is only the \texttt{JUMP\_IF\_A}
instruction, and only if it is actually a backward jump. The place where this
method is called is where the language interpreter performs profiling to decide
when to start tracing. It is also the place where the tracing JIT checks
whether a loop is closed. This is considered to be the case when the values of
the "green" variables are the same as at an earlier call to the
\texttt{can\_enter\_jit} method.

For the small example the hints look like a lot of work. However, the number of
hints is essentially constant no matter how large the interpreter is, which
makes it seem less significant for larger interpreters.

@@

The critical problem of tracing the execution of just one opcode has been
solved, the loop corresponds exactly to the loop in the square function.
However, the resulting trace is a bit too long. Most of its operations are not
actually doing any computation that is part of the square function. Instead,
they manipulate the data structures of the language interpreter. While this is
to be expected, given that the tracing interpreter looks at the execution of the
language interpreter, it would still be an improvement if some of these operations could
be removed.

@@

With this optimization enabled, the trace looks as in Figure
\ref{fig:trace-full}. Now much of the language interpreter is actually gone
from the trace and what is left corresponds very closely to the loop of the
square function. The only vestige of the language interpreter is the fact that
the register list is still used to store the state of the computation. This
could be removed by some other optimization, but is maybe not really all that
bad anyway (in fact we have an experimental optimization that does exactly that,
but it is not yet finished).  Once we get this optimized trace, we can pass it to
the \emph{JIT backend}, which generates the correspondent machine code.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It should be possible to choose when the language interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
that when no hints were present in the interpreter at all.

@@

What is done instead is that the language interpreter keeps running as a C
program, until a hot loop in the user program is found. To identify loops, the
C version of the language interpreter is generated in such a way that at the
place that corresponds to the \texttt{can\_enter\_jit} hint profiling is
performed using the program counter of the language interpreter. Apart from this
bit of profiling, the language interpreter behaves in just the same way as
without a JIT.

@@

This section will look at some other implementation issues and optimizations
that we have done that are beyond the scope of this paper (and will be the subject
of a later publication).

@@

\textbf{Allocation Removal:} A key optimization for making the approach
produce good code for more complex dynamic language is to perform escape
analysis on the loop operation after tracing has been performed. In this way all
objects that are allocated during the loop and do not actually escape the loop do
not need to be allocated on the heap at all but can be exploded into their
respective fields.  This is very helpful for dynamic languages where primitive
types are often boxed, as the constant allocation of intermediate results is
very costly.

@@

In this section we try to evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1 GB RAM, using Linux 2.6.27. All benchmarks where
run 50 times, each in a newly started process. The first run was ignored. The
final numbers were reached by computing the average of all other runs, the
confidence intervals were computed using a 95\% confidence level. All times
include the running of the tracer and machine code production to measure how
costly those are.

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
the square of 46340 (the smallest number whose square still fits into a 32
bit word) using the bytecode of Figure \ref{fig:square}. The results for various
constellations are as follows:

@@

\textbf{Benchmark 4:} Same as before, but with constant folding enabled. This corresponds to the
trace in Figure \ref{fig:trace-full}. This speeds up the square function considerably,
making it about six times faster than the pure interpreter.

@@

Classical partial evaluation has failed to be useful for dynamic language for
much the same reasons why ahead-of-time compilers cannot compile them to
efficient code. If the partial evaluator knows only the program it simply does
not have enough information to produce good code. Therefore some work has been
done to do partial evaluation at runtime. One of the earliest works on runtime
specialisation is Tempo for C \cite{consel_general_1996, consel_uniform_1996}.
However, it is essentially a normal
partial evaluator ``packaged as a library''; decisions about what can be
specialised and how, are pre-determined. Another work in this direction is DyC
\cite{grant_dyc:expressive_2000}, another runtime specializer for C. Both of these projects
have a problem similar to that of DynamoRIO.  Targeting the C language makes
higher-level specialisation difficult (e.g.\ \texttt{malloc} cannot be
optimized).

There have been some attempts to do \emph{dynamic partial evaluation}, which is
partial evaluation that defers partial evaluation completely to runtime
to make partial evaluation more useful for dynamic languages. This concept was
introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
small dynamic language based on lambda-calculus. There is some work by one of
the authors to implement a dynamic partial evaluator for Prolog
\cite{carl_friedrich_bolz_automatic_2008}. There are also experiments within the
PyPy project to use dynamic partial evaluation for automatically generating JIT
compilers out of interpreters \cite{armin_rigo_jit_2007}. So far those have not been as
successful as we would like and it seems likely that they will be supplanted
with the work on tracing JITs described here.

@@

We have shown techniques for improving the results when applying a tracing
JIT to an interpreter. Our first benchmarks indicate that these techniques work
really well on small interpreters and first experiments with PyPy's Python
interpreter make it appear likely that they can be scaled up to realistic
examples.

Of course there is a lot of work still left to do. Various optimizations are not
quite finished. Both tracing and leaving machine code is very slow due to a
double interpretation overhead and we might need techniques for improving those.
Furthermore we need to apply the JIT to the various interpreters that are
written in RPython to evaluate how widely applicable the described techniques
are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
implementation \cite{bolz_back_2008}, a Prolog interpreter or PyGirl, a Gameboy
emulator \cite{XXX}; but also less immediately obvious ones, like Python's
regular expression engine.
